---
title: "SARS-CoV-2 in dogs and cats"
output: html_document
date: '2022-07-14'
---

# Load packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(janitor)
library(ggplot2)
library(readxl)
library(gdata)
library(tidyverse)
library(openxlsx)
library(data.table)
library(cowplot)
library(tiff)
library(stringr)
library(broom)
library(epitools)
library(ggpubr)
library(DescTools)
setwd("/Users/tegin/Desktop/GitHub2/SARS-CoV-2_in_housepets/R_files_final")
```

# Read-in spreadsheets

```{r}
sample_data<-read.xls("sample_data.xlsx", header = TRUE)

summary(sample_data)
sample_data$Species<-as.factor(sample_data$Species)
sample_data$Timepoint<-as.factor(sample_data$Timepoint)
sample_data$Sample_Type<-as.factor(sample_data$Sample_Type)
sample_data$Denny_run<-as.factor(sample_data$Denny_run)
sample_data$Denny_result<-as.factor(sample_data$Denny_result)
sample_data$COBAS_result<-as.factor(sample_data$COBAS_result)
sample_data$Control_real<-as.factor(sample_data$Control_real)
sample_data$IgG_result<-as.factor(sample_data$IgG_result)
sample_data$IgM_resulte<-as.factor(sample_data$IgM_result)
sample_data$VN_result<-as.factor(sample_data$VN_result)

patient_data <- read.xls("patient_data.xlsx", header = TRUE)
patient_data<-patient_data %>% 
  mutate(across('hp_species', str_replace, '1', 'Canine')) %>% 
  mutate(across('hp_species', str_replace, '2', 'Feline'))

summary(patient_data)
patient_data$hp_species<-as.factor(patient_data$hp_species)
patient_data$hp_breed<-as.factor(patient_data$hp_breed)
patient_data$hp_sex<-as.factor(patient_data$hp_sex)
patient_data$hp_neuter_yn_unk<-as.factor(patient_data$hp_neuter_yn_unk)
patient_data$exposed_yn<-as.factor(patient_data$exposed_yn)
patient_data$hp_medical_hx<-as.factor(patient_data$hp_medical_hx)
patient_data$hp_medications<-as.factor(patient_data$hp_medications)
patient_data$area_class<-as.factor(patient_data$area_class)
patient_data$contact_bed<-as.factor(patient_data$contact_bed)
patient_data$contact_furn<-as.factor(patient_data$contact_furn)
patient_data$contact_hands<-as.factor(patient_data$contact_hands)
patient_data$contact_face<-as.factor(patient_data$contact_face)
patient_data$contact_body<-as.factor(patient_data$contact_body)
patient_data$eat_habit_humandish<-as.factor(patient_data$eat_habit_humandish)
patient_data$contact_humandish<-as.factor(patient_data$contact_humandish)
patient_data$contact_wildanimals<-as.factor(patient_data$contact_wildanimals)
patient_data$excretehome_habits<-as.factor(patient_data$excretehome_habits)
patient_data$bites_yn<-as.factor(patient_data$bites_yn)
patient_data$scratch_yn<-as.factor(patient_data$scratch_yn)
patient_data$hp_sx_present<-as.factor(patient_data$hp_sx_present)
patient_data$hp_sx_lethargy<-as.factor(patient_data$hp_sx_lethargy)
patient_data$hp_sx_anorexia<-as.factor(patient_data$hp_sx_anorexia)
patient_data$hp_sx_fever<-as.factor(patient_data$hp_sx_fever)
patient_data$hp_sx_nausea<-as.factor(patient_data$hp_sx_nausea)
patient_data$hp_sx_diarrhea<-as.factor(patient_data$hp_sx_diarrhea)
patient_data$hp_sx_sneezing<-as.factor(patient_data$hp_sx_sneezing)
patient_data$hp_sx_coughing<-as.factor(patient_data$hp_sx_coughing)
patient_data$hp_sx_itchy_watery_eyes<-as.factor(patient_data$hp_sx_itchy_watery_eyes)
patient_data$hp_sx_runny_nose<-as.factor(patient_data$hp_sx_runny_nose)
patient_data$hp_sx_gums<-as.factor(patient_data$hp_sx_gums)
patient_data$hp_sx_retching<-as.factor(patient_data$hp_sx_retching)
patient_data$hp_sx_breath<-as.factor(patient_data$hp_sx_breath)
patient_data$hp_sx_nm_eyelid<-as.factor(patient_data$hp_sx_nm_eyelid)
patient_data$hp_sx_other<-as.factor(patient_data$hp_sx_other)
patient_data$Denny_result<-as.factor(patient_data$Denny_result)
patient_data$COBAS_result<-as.factor(patient_data$COBAS_result)
patient_data$IgG_result<-as.factor(patient_data$IgG_result)
patient_data$IgM_result<-as.factor(patient_data$IgM_result)
patient_data$Any_seropos<-as.factor(patient_data$Any_seropos)
patient_data$Virus_neutralization<-as.factor(patient_data$Virus_neutralization)
```

# Sample types collected by house and day, possible figure

```{r}
sample_data$Paper_household_ID<-as.factor(sample_data$Paper_household_ID)
sample_data %>% 
  drop_na(Timepoint) %>% 
  mutate(Timepoint=factor(Timepoint, levels=c("0", "1", "3", "7", "14", "21", "28", "45", "60", "120", "180"))) %>%
  mutate(Paper_househould_ID=factor(Paper_household_ID, levels=c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31", "32"))) %>% 
ggplot() + 
  ggtitle("Sample Types Collected By House and Day")+
  xlab("Timepoint (Day)") + ylab("House Number") +
  geom_point(mapping = aes(x = Timepoint, y = Paper_household_ID, color=Household_ID)) +
  theme_bw() +
  theme(plot.title = element_text(hjust=0.5), axis.text.y = element_text(angle=0, size=5), legend.position="None", axis.text.x = element_text(angle=30, size=5))+
  facet_wrap(~Sample_Type)+
  scale_y_discrete(guide = guide_axis(n.dodge=2))
```
# Figure 1

## Sample types collected by species
```{r}
figure_1<-sample_data %>% 
  ggplot() +
  geom_bar(aes(x= Species, fill= Sample_Type), position = position_dodge(preserve = 'single')) +
  ggtitle("Sample Types Collected By Species")+
  xlab("Species sampled") +
  ylab("Number of samples") +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(fill=guide_legend(title="Sample Type"))



ggsave("/Users/tegin/Desktop/COVID_manuscript/manuscript/Submission_3/figure_1.pdf", plot=figure_1)


```

# Figure 2

## Dog IgG ODs

```{r}
D_IgG<-
  sample_data %>% 
  filter(Species=="Canine") %>% 
  ggplot()+
  geom_point(aes(x=IgG1_OD, y=IgG2_OD, color=Control_real))+
  xlim(0, 2) +
  ylim(0, 2.5) +
  theme_bw() +
  geom_point(data=subset(sample_data, Control_real=="Pre-2019 serological control" & Species=="Canine"),mapping = aes(x=IgG2_OD, y=IgG2_OD))+
  labs(color = "Sample Type", tag='A')+
       geom_point(data=sample_data %>% filter(IgG1_OD==0.69),
             pch=21,
             fill=NA,
             size=4,
             color="black", mapping=aes(x=IgG2_OD, y=IgG2_OD)) +
  theme(plot.title = element_text(hjust = 0.5), 
        legend.text=element_text(size=12),
        legend.title = element_blank()) +
  scale_color_manual(values = c("Pre-2019 serological control" = "black", "Exposed serum sample" = "coral"))+
  geom_vline(xintercept = 1, linetype="dotted", 
                color = "black", size=0.5) +
  geom_hline(yintercept = 1, linetype="dotted", 
                color = "black", size=0.5)+
  geom_text(aes(1, 0.9, label = "Proposed cut-off=1.0", hjust = -.1), check_overlap = TRUE) +
  geom_text(aes(0.38, 1.77, label = "Outlier", hjust = -.1), check_overlap = TRUE) +
  ggtitle("Dog IgG") +
  xlab("Round 1") + ylab("Round 2") 
```

## Dog IgM ODs

```{r}
D_IgM<-sample_data %>% 
  filter(Species=="Canine") %>% 
  ggplot()+
  geom_point(aes(x=IgM1_OD, y=IgM2_OD, color=Control_real))+
  xlim(0, 2) +
  ylim(0, 2.5) +
  theme_bw() +
  geom_point(data=subset(sample_data, Control_real=="Pre-2019 serological control" & Species=="Canine"),mapping = aes(x=IgM2_OD, y=IgM2_OD))+
  labs(color = "Sample Type", tag='B')+
       geom_point(data=sample_data %>% filter(IgM2_OD==0.525),
             pch=21,
             fill=NA,
             size=4,
             color="black", mapping=aes(x=IgM2_OD, y=IgM2_OD)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("Pre-2019 serological control" = "black", "Exposed serum sample" = "coral"))+
  #geom_vline(xintercept = 0.75, linetype="dotted", 
               # color = "black", size=0.5) +
 # geom_hline(yintercept = 0.75, linetype="dotted", 
               # color = "black", size=0.5)+
  #geom_text(aes(.73, 0.65, label = "Proposed cut-off=0.75", hjust = -.1), check_overlap = TRUE) +
  geom_text(aes(0.15, 2.3, label = "Outlier", hjust = -.1), check_overlap = TRUE) +
  ggtitle("Dog IgM") +
  xlab("Round 1") + ylab("Round 2") 
```

## Cat IgG ODs

```{r}
C_IgG<-
  sample_data %>% 
  filter(Species=="Feline") %>% 
  ggplot()+
  geom_point(aes(x=IgG1_OD, y=IgG2_OD, color=Control_real))+
  xlim(0, 2) +
  ylim(0, 2.5) +
  theme_bw() +
  geom_point(data=subset(sample_data, Control_real=="Pre-2019 serological control" & Species=="Feline"),mapping = aes(x=IgG2_OD, y=IgG2_OD))+
  labs(color = "Sample Type", tag='C')+
       geom_point(data=sample_data %>% filter(IgG1_OD==0.164 | IgG1_OD==0.18 | IgG1_OD==0.96),
             pch=21,
             fill=NA,
             size=4,
             color="black", mapping=aes(x=IgG2_OD, y=IgG2_OD)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("Pre-2019 serological control" = "black", "Exposed serum sample" = "coral"))+
  geom_vline(xintercept = 0.8, linetype="dotted", 
                color = "black", size=0.5) +
  geom_hline(yintercept = 0.8, linetype="dotted", 
                color = "black", size=0.5)+
  geom_text(aes(0.8, 0.70, label = "Proposed cut-off=0.8", hjust = -.1), check_overlap = TRUE) +
  ggtitle("Cat IgG") +
  xlab("Round 1") + ylab("Round 2") 
```

## Cat IgM ODs

```{r}

C_IgM<-sample_data %>% 
  filter(Species=="Feline") %>% 
  ggplot()+
  geom_point(aes(x=IgM1_OD, y=IgM2_OD, color=Control_real))+
  xlim(0, 2) +
  ylim(0, 2.5) +
  theme_bw() +
  geom_point(data=subset(sample_data, Control_real=="Pre-2019 serological control" & Species=="Feline"),mapping = aes(x=IgM2_OD, y=IgM2_OD))+
  labs(color = "Sample Type", tag='D')+
  scale_color_manual(values = c("Pre-2019 serological control" = "black", "Exposed serum sample" = "coral"))+
  geom_point(data=sample_data %>% filter(IgM2_OD==0.79 | IgM2_OD==0.767 | IgM2_OD==0.494),
             pch=21,
             fill=NA,
             size=4,
             color="black", mapping=aes(x=IgM2_OD, y=IgM2_OD)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Cat IgM") +
  xlab("Round 1") + ylab("Round 2") 
```

## Combine into one figure

```{r}
OD_figure <- ggarrange(D_IgG, D_IgM, C_IgG, C_IgM, ncol=2, nrow=2, common.legend = TRUE, legend='bottom', font.label = list(size=24))
OD_figure <- annotate_figure(OD_figure)#, 
                            # top = text_grob("Round 1 vs Round 2 Optical Densities for Canine and Feline anti-SARS-CoV-2 IgG and IgM", 
                            # face = "bold",
                             #size = 16))


OD_figure
#ggsave("Figure_1.pdf", plot=OD_figure, width=35, height=20, units="cm")

ggsave("/Users/tegin/Desktop/COVID_manuscript/manuscript/Submission_3/figure_2.pdf", plot=OD_figure, width=35, height=20, units="cm")
```


# Demographic data

## Species, age, breed, household breakdowns

```{r}
sample_data %>% 
  filter(Control_real!="Pre-2019 control") %>% 
  distinct(Subject_ID, Species) %>% 
  count(Species)

patient_data %>% 
  count(hp_breed, sort=TRUE)

summary(patient_data$hp_age)

patient_data %>% 
  group_by(hp_species) %>% 
  summarise(median(hp_age, na.rm=TRUE), range(hp_age, na.rm=TRUE))
```

# Sample collection data

## PCR

```{r}
table(sample_data$Sample_Type, sample_data$Species)

table(sample_data$Sample_Type, sample_data$Timepoint)

sample_data %>% 
  filter(Sample_Type!="Serum") %>% 
  select(Subject_ID, Species, Timepoint) %>% 
  group_by(Subject_ID) %>% 
  summarize(n=n_distinct(Timepoint), Species) %>% 
  filter(n>1) %>% 
  distinct(Subject_ID, Species) %>% 
  ungroup(Subject_ID) %>% 
  count(Species)
```

## Serum

```{r}
table(sample_data$Sample_Type=="Serum", sample_data$Control_real)

sample_data %>% 
  filter(Control_real=="Real sample" & IgG_result!="Outlier") %>% 
  count(Species)
```

# Positive sample data

## PCR, table for PCR positive animals

```{r}
Table_1<-sample_data %>% 
  filter(Denny_result=="Positive") %>% 
  select(Paper_household_ID, Species, Timepoint, Sample_Type, VL_RNA_cp_per_mL, Denny_result, COBAS_result) %>% 
  rename("Household"=Paper_household_ID, "Sample"=Sample_Type, "Viral Load (RNA cp/mL)"=VL_RNA_cp_per_mL, "PCR Result"=Denny_result, "COBAS result"=COBAS_result)
write_csv(Table_1, "Table_1.csv")
```

## Serology results

```{r}
patient_data %>% 
  filter(Any_seropos!="NA", IgG_result!="Outlier") %>% 
  select(hp_species, IgG_result, IgM_result) %>% 
  count(hp_species, IgG_result, IgM_result) %>% 
  rename("Species"=hp_species, "IgG Result"=IgG_result, "IgM Result"=IgM_result, "Number Positive"=n)

Table_2<-sample_data %>% 
  filter(IgG_result!="NA" & IgG_result!="Outlier" & IgG_result!="") %>% 
  filter(IgG_result=="Positive" | IgM_result=="Positive") %>% 
  select(Paper_household_ID, Species, Timepoint, IgM_result, IgG_result) %>%
  rename("House Number"=Paper_household_ID, "IgG Result"=IgG_result, "IgM Result"=IgM_result, "Timepoint (Days)"=Timepoint)

write_csv(Table_2, "Table 2.csv")

patient_data %>% 
  filter(Paper_house_ID=="5" | Paper_house_ID=="22")
```

# Stats

## Calculate Lin's correlation coefficient for OD runs
```{r}
dog_sample_data <- sample_data %>% 
  filter(Species=="Canine")

cat_sample_data <- sample_data %>% 
  filter(Species=="Feline")

CCC(sample_data$IgG1_OD, sample_data$IgG2_OD, ci="z-transform", conf.level=0.95, na.rm=TRUE)
CCC(sample_data$IgM1_OD, sample_data$IgM2_OD, ci="z-transform", conf.level=0.95, na.rm=TRUE)
CCC(dog_sample_data$IgG1_OD, dog_sample_data$IgG2_OD, ci="z-transform", conf.level=0.95, na.rm=TRUE)
CCC(dog_sample_data$IgM1_OD, dog_sample_data$IgM2_OD, ci="z-transform", conf.level=0.95, na.rm=TRUE)
CCC(cat_sample_data$IgG1_OD, cat_sample_data$IgG2_OD, ci="z-transform", conf.level=0.95, na.rm=TRUE)
CCC(cat_sample_data$IgM1_OD, cat_sample_data$IgM2_OD, ci="z-transform", conf.level=0.95, na.rm=TRUE)
```


## Separate dog and cat

```{r}
dog_patient_data <-
  patient_data %>% 
  filter(hp_species=="Canine" & IgG_result!="Outlier")

cat_patient_data <-
  patient_data %>% 
  filter(hp_species=="Feline")
```

## Create functions for fisher.test and ORs
```{r}
fisher_pval <- function(x, y){
  fisher.test(x, y)$p.value
}

odds_ratio = function(x, y){
  a = table(x, y)[[1, 1]]
  b = table(x, y)[[1, 2]]
  c = table(x, y)[[2, 1]]
  d = table(x, y)[[2, 2]]
  #Apply Haldane-Anscombe correction if necessary
  if(a*b*c*d == 0){
    OR = ((a+0.5) * (d+0.5)) / ((b+0.5) * (c+0.5))
  } else {
    OR = (a*d) / (b*c)
  }
}

test_columns <- function(df, cols, comp_col, test){
  res <- df %>% 
    select(all_of(cols)) %>% 
    sapply(function(x) test(droplevels(df[[comp_col]]), x))
}
```

## Fisher.test, ORs, p-value tables
```{r}
fisher_cols <- c('hp_sex', 'hp_neuter_yn_unk', 'area_class', 'contact_bed', 'contact_furn', 'contact_hands', 'contact_face', 'contact_body', 'eat_habit_humandish', 'contact_humandish', 'contact_wildanimals', 'excretehome_habits', 'bites_yn', 'scratch_yn')

dog_and_cat_data <- patient_data %>% 
  filter(IgG_result!="Outlier")

## Dog/cat any seropos
p_val_dog_cat_any<- test_columns(dog_and_cat_data, fisher_cols, 'Any_seropos', fisher_pval)

## Dog any
p_val_dog_any <- test_columns(dog_patient_data, fisher_cols, 'Any_seropos', fisher_pval)

## Dog IgG
p_val_dog_IgG <- test_columns(dog_patient_data, fisher_cols, 'IgG_result', fisher_pval)

## Dog IgM
p_val_dog_IgM <- test_columns(dog_patient_data, fisher_cols, 'IgM_result', fisher_pval)

## Cat IgG
p_val_cat_IgG <-  test_columns(cat_patient_data, fisher_cols, 'IgG_result', fisher_pval)

p_val<-cbind(dog_cat_any=p_val_dog_cat_any,
             dog_any=p_val_dog_any,
             dog_IgG=p_val_dog_IgG,
             dog_IgM=p_val_dog_IgM,
             cat_IgG=p_val_cat_IgG)
p_val<-as.data.frame(p_val)

p_val

sig_p_val<-p_val %>% 
  filter(dog_cat_any<.05 | dog_any<.05 | dog_IgG<.05 | dog_IgM<.05 | cat_IgG<.05)

adjusted_p<-p.adjust(p_val_dog_IgG, method="bonferroni")

## Dog/cat any seropos
OR_dog_cat_any<- test_columns(dog_and_cat_data, fisher_cols, 'Any_seropos', odds_ratio)

## Dog any
OR_dog_any <- test_columns(dog_patient_data, fisher_cols, 'Any_seropos', odds_ratio)

## Dog IgG
OR_dog_IgG <- test_columns(dog_patient_data, fisher_cols, 'IgG_result', odds_ratio)

## Dog IgM
OR_dog_IgM <- test_columns(dog_patient_data, fisher_cols, 'IgM_result', odds_ratio)

## Cat IgG
OR_cat_IgG <-  test_columns(cat_patient_data, fisher_cols, 'IgG_result', odds_ratio)

OR <- cbind(dog_cat_any=OR_dog_cat_any,
            dog_any=OR_dog_any,
            dog_IgG=OR_dog_IgG,
            dog_IgM=OR_dog_IgM,
            cat_IgG=OR_cat_IgG)
OR <- as.data.frame(OR)
OR
```
## Symptom data
```{r}
summary(patient_data$hp_sx_present)
table(patient_data$hp_sx_present, patient_data$Denny_result)
table(patient_data$hp_sx_present, patient_data$IgG_result)
patient_data %>% 
  filter(hp_sx_present==1, hp_species=="Canine", IgG_result=="Positive")
fisher.test(patient_data$hp_sx_present, patient_data$IgG_result)
```


## p-value plots

```{r}
p_val %>% 
  ggplot(aes(x=dog_cat_any)) +
  geom_histogram(colour="black", fill="white", binwidth=0.05)

p_val %>% 
  ggplot(aes(x=dog_any)) +
  geom_histogram(colour="black", fill="white", binwidth=0.05)

p_val %>% 
  ggplot(aes(x=dog_IgG)) +
  geom_histogram(colour="black", fill="white", binwidth=0.05)

p_val %>% 
  ggplot(aes(x=dog_IgM)) +
  geom_histogram(colour="black", fill="white", binwidth=0.05)

p_val %>% 
  ggplot(aes(x=cat_IgG)) +
  geom_histogram(colour="black", fill="white", binwidth=0.05)
```


```{r}
sample_data %>% 
  filter(Species=="Canine") %>% 
  select(Subject_ID, IgG_result, Denny_result, Timepoint)
```

